---
title: "Differential Expression of Rare  Variants"
author: "Jenny Smith"
date: "June 20, 2017"
output: html_document
---


```{r setup}
library(knitr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE, fig.align='center', fig.height = 10, fig.width = 10)
knitr::opts_knit$set(root.dir = '~/RNA_seq_Analysis/2017.10.25_1031_NUP98-NSD1/')
```


```{r message = FALSE, warning=FALSE}
library(stringr)
library(magrittr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(tibble)
getwd()
```


```{r}
source("~/scripts/RNAseq_Analysis/DifferentialExpn_PathwayAnalysis/DifferentialExpressionPipeline.r")
source("~/scripts/survival_analysis/Survplot_Functions_2017.03.31.r")
source("~/scripts/RNAseq_Analysis/MSLN_waterfallplots/Waterfall_Barplot_Function_2017.05.22.r")
```




#Define Functions to be used


```{r}
mutationGroup <- function(NPM1,CEBPA,NUP98.NSD1,CBF=FALSE,inv.16.16_t.16.16=NULL,t.8.21=NULL){
#yes,this assumes if CBF-AML + FLT3-ITD, does not have any additional co-occuring mutations. 
  
  NPM1orCEBPA <-  any(grepl("Pos|Yes",c(NPM1,CEBPA),ignore.case=TRUE))
  NSD1 <- grepl("Pos|Yes", NUP98.NSD1, ignore.case = TRUE)
  
  if (NPM1orCEBPA){
    grp <- "NPM1 or CEBPA, FLT3-ITD"

  }else if (NSD1){
    grp <- "NUP98-NSD1, FLT3-ITD" #NOTE THERE IS ONE NUP98-NSD1 + NPM1 + FLT3-ITD patient.... in0531

  }else{
    grp <- "FLT3-ITD"

  }

  if (CBF){ #ONE patient is NSD1+ and Inv16 ... in 0531
    grp[grepl("Pos|Yes", inv.16.16_t.16.16, ignore.case = TRUE) | grepl("Pos|Yes",t.8.21, ignore.case = TRUE)] <- "CBF, FLT3-ITD"
  }

  return(grp)
}
```



#Read in the counts

```{r}
cts.1031 <- read.csv("~/RNA_seq_Analysis/0000.00.03_Expression_Matrices/TARGET_AML_AAML1031_dupGenesRemoved_FractionalCounts.csv", stringsAsFactors = FALSE, row.names = 1)
colnames(cts.1031) <- str_split_fixed(colnames(cts.1031),  "\\.", 2)[,1]
head(cts.1031[,1:5])
```

```{r}
cts.0531 <- read.csv("~/RNA_seq_Analysis/0000.00.03_Expression_Matrices/TARGET_AML_AAML0531_dupGenesRemoved_fractionalCounts.csv", stringsAsFactors = FALSE,
                     row.names = 1)
head(cts.0531[,1:5])
```

```{r}
cts.0531.HD <- read.csv("~/RNA_seq_Analysis/0000.00.03_Expression_Matrices/TARGET_AML_DxSamples_dupGenesRemoved_cts_Aug2014.csv", stringsAsFactors = FALSE, row.names = 1)

colnames(cts.0531.HD) <- str_split_fixed(colnames(cts.0531.HD), "\\.", n=5)[,3]
head(cts.0531.HD[,1:5])
```


```{r}
tpm <- read.csv("ExpressionData/TARGET_AML_AAML1031_dupGenesRemoved_TPM.csv", stringsAsFactors = FALSE,
                row.names = 1)

head(tpm[,1:5])
```




#Read in the clinical Data


```{r}
CDE.1031 <- read.csv("~/reference_mapping-files/TARGET_AML_1031_CDE_cleaned_3.20.18.csv",
                     stringsAsFactors = FALSE)

CDE.1031 <- CDE.1031 %>%
  filter(!is.na(USI)) %>%
  column_to_rownames("USI")

head(CDE.1031)
dim(CDE.1031)
```

```{r}
table(CDE.1031$NUP98.NSD1)
```

```{r}
# FLT3.NSD1 <- read.csv("AAML1031_Combined_CDE_ITDpos_Cleaned_10.25.2017.csv", stringsAsFactors = FALSE, row.names = 1)
FLT3.NSD1 <- read.csv("ITDpos_Cleaned_withMolecularAdded_10.24.2017.csv", stringsAsFactors = FALSE)
head(FLT3.NSD1[,1:5])
```

```{r}
CDE.0531 <- read.csv("~/reference_mapping-files/TARGET_AML_current_asof_june30_2016_UPDATED_CLEAN_3.27.18.csv", stringsAsFactors = FALSE, row.names = 1)
head(CDE.0531[,1:5])
```


```{r}
table(CDE.0531$NUP98.NSD1)
```


```{r}

```


#Perform Differential Expression Analysis 

Limma Trend with LogCPM and TMM normalized counts. Rare Variant AML vs All Other AML

Using Fractional Counts as Input

```{r}
merged <- read.csv("~/reference_mapping-files/TARGET_AML_1031_0531_Merged_CDE_3.22.18.csv", stringsAsFactors = FALSE)

merged <- merged %>% 
  filter(!is.na(TARGET.USI.1)) %>%
  set_rownames(.$TARGET.USI.1)

# dim(merged)
head(merged[,1:5])
```

```{r}
table(merged$NUP98.NSD1)
```

```{r}
cts.list <- list("HD.0531"=cts.0531.HD, "LD.0531"=cts.0531,"HD.0531"=)
```


```{r message=FALSE,warning=FALSE}
# tiff(file="TARGET_AML_NUP98-NSD1_vs_otherAML_DEGs_Heatmaps_2.27.17.pdf", height = 10, width = 10,
     # units="in", res=600)
DEGs.NSD1 <- twoGroups_DEGs(expnData=cts.1031, 
                                    clinData=CDE.1031, 
                                    col="NUP98.NSD1",
                                    ref="No",
                                    logCPM = TRUE)

# dev.off()
```


```{r}
summary(DEGs.NSD1)
```


```{r}
DEGs.df <- extract_DEGs(DEGs.NSD1)


# dim(DEGs.df) #3,103 DEGs
head(DEGs.df)
```


```{r}
# write.csv(DEGs.df, "TARGET_AML_1031_NUP98-NSD1_vs_OtherAML_DEGS_2.27.18.csv", row.names = FALSE)
```


#Examine QC plots

```{r}
# pdf("NUP98-NSD1vsOtherAML_MDS.pdf")
extract_MDS(DEGs.NSD1)
# dev.off()
```


```{r}
# pdf("NUP98-NSD1vsOtherAML_PCA.pdf")
extract_PCA(DEGs.NSD1)
# dev.off()
```



#Differential Expression of AML Rare Variants versus Normal BM

```{r}
# pdf(file="TARGET_AML_NUP98-NSD1_vs_NormalBM_DEGs_Heatmaps_25Oct2017.pdf")
DEGs.NSD1vBM <- twoGroups_DEGs(expnData=cts.1031,
                               clinData=NSD1,
                               col="NUP98.NSD1",
                               ref="negative",
                               logCPM=TRUE,
                               BM=TRUE)

# dev.off()
```

```{r}
summary(DEGs.NSD1vBM)
```

```{r}
NBM.df <- extract_DEGs(DEGs.NSD1vBM)
head(NBM.df)
```


```{r}
# write.csv(NBM.df, "TARGET_AML_1031_NUP98-NSD1_vs_NormalBM_DEGs.csv", row.names = FALSE)
```

#Check out the QC plots

```{r}
# pdf("NUP98-NSD1_vs_NormalBM_PCA.pdf")
extract_PCA(DEGs.NSD1vBM)
# dev.off()
```


```{r}
# pdf("NUP98-NSD1_vs_NormalBM_MSD.pdf")
extract_MDS(DEGs.NSD1vBM)
# dev.off()
```


#Differential Expression of FLT3-ITD with NUP98-NSD1 vs FLT3-ITD Alone

mutate(ITD.NSD1=ifelse(grepl("FLT3-ITD", MutGrps4), "ITD", "ITD.NSD1")) %>%
```{r}
# idx <- FLT3.NSD1$MutGrps4 != "CBF, FLT3-ITD" & FLT3.NSD1$MutGrps4 != "NPM1 or CEBPA, FLT3-ITD"
FLT3alone <- FLT3.NSD1 %>%
  filter(MutGrps4 != "CBF, FLT3-ITD" & MutGrps4 != "NPM1 or CEBPA, FLT3-ITD" & USI != "#N/A") %>%
  mutate(ITD.NSD1=ifelse(grepl("NSD1", MutGrps4), "FLT3NSD1", "FLT3ITD")) %>%
  set_rownames(.$USI)

dim(FLT3alone)
```

NOTE: THERE IS AN AWEFUL BUG where some combination of character levels leads to odd duplication of USIs??
happenw with ITD and ITD.NSD1 as levels...

```{r}
idx <- rownames(FLT3alone) %in% colnames(cts.1031)
table(FLT3alone$ITD.NSD1[idx])
```


```{r}
# pdf("TARGET_AML_1031_FLT3withNUP98-NSD1_vs_FLT3alone_Heatmap.pdf")
DEGs.FLT3 <- twoGroups_DEGs(expnData = cts.1031, 
                            clinData = FLT3alone,
                            col = "ITD.NSD1", 
                            ref = "FLT3ITD", 
                            logCPM = TRUE)

# dev.off()
```


```{r}
summary(DEGs.FLT3)
```

```{r}
# table(DEGs.FLT3)
```

```{r}
FLT3.df <- extract_DEGs(DEGs.FLT3)
head(FLT3.df)
```

```{r}
# write.csv(FLT3.df, "TARGET_AML_1031_FLT3withNUP98-NSD1_vs_FLT3alone_DEGs.csv", row.names = FALSE)
```


```{r}
extract_MDS(DEGs.FLT3)
```

```{r}
# pdf("LT3withNUP98-NSD1_vs_FLT3alone_PCA.pdf")
extract_PCA(DEGs.FLT3)
# dev.off()
```


#Identify NUP98.NSD1-like genes 


```{r}
source("~/scripts/RNAseq_Analysis/DifferentialExpn_PathwayAnalysis/corrplot_Function.r")
```


```{r}
  #find absolute correlation values 
  gene.cor.abs <- abs(cor(t(cts.1031[DEGs.df$gene,])))
  gene.cor.abs[1:10, 1:10]
```

```{r}
  #function to pullout the highest genes correlation pair
  maxAbsCor <- function(x){
    x <- x[x < 1] #remove the correlation to itself
    x <- x[x == max(x)]
    res <- data.frame(cor=x,
                      pair=names(x), 
                      stringsAsFactors = FALSE)
    
    return(res)
  }
```

```{r}
  #apply the function to each row or column (doesn't matter its a diagnol upper and lower matrix still)
maxcors <- bind_rows(apply(gene.cor.abs,2, maxAbsCor)) %>%
    mutate(gene=colnames(gene.cor.abs)) %>%
    arrange(desc(cor)) %>%
    dplyr::slice(seq(1,nrow(.), by=2)) %>% #remove recipricol pairs 
    filter(cor >= quantile(cor, c(0.95))) #select correlations in the 80th percentile or greater 

head(maxcors)
dim(maxcors)
```


```{r}
corrPairs <- maxCorrPairs(cts.1031[DEGs.df$gene,])
```


```{r}
DEGs.df.sig <- DEGs.df %>%
  filter(adj.P.Val < 0.0001) %>%
  filter(gene %in% maxcors$pair | gene %in% maxcors$gene)

dim(DEGs.df.sig)
```


```{r}
d.top <- dge_dendrograms(expData = DEGs.NSD1$InputExpnMatrix, pheno = DEGs.NSD1$phenovector,   )
```





#Waterfall plot of Genes of Interest


```{r}
ph <- ifelse(grepl("GroupA", DEGs.FLT3$phenovector ), "FLT3.NSD1", "FLT3.alone") %>% 
  set_names(names(DEGs.FLT3$phenovector))

NBMorOtherAML <- ifelse(grepl("^BM|^RO",colnames(tpm)), "NBM", "otherAML")%>%
  set_names(colnames(tpm)) %>%
  subset(.,! names(.) %in% names(ph))

pheno <- c(ph,NBMorOtherAML) %>% set_names(c(names(ph), names(NBMorOtherAML)))

ph[1:10]
pheno[c(1:5,1000,1100)]
```

```{r}
table(pheno)
```


```{r}
waterfallPlot(tpm,"CSF1R",pheno,unit="TPM")
```

```{r}
df.csf1r <- data.frame(t(tpm), 
                  Patient=names(tpm),
                  Status=phenoVector)
```



#Look at the 0531 Data


```{r}
# FLT3.0531 <- CDE.0531 %>%

```



#Session Information 

```{r}
sessionInfo()
```



